Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-01-03 supplement.tex
removed speculation
refit LLs

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/supp-merged-data-model-fits}
    \caption{
        \textbf{Merged datasets bias demographic inferences.}
        The panels represent best-fit models according to 
        different combinations of high- and low- coverage data.
        While statistics from low and high coverage datasets provide consistent
        best-fit demographic models, merged datasets result in biased LD
        statistics (Figure~\ref{fig:supp-merged-data-SEs}). These biased LD
        statistics lead to large downstream biases in inferred demographic
        models. 
    }
    \label{fig:supp-merged-data-fits}
\end{figure}

@@

>>>>

\begin{figure}[htb!]
    \centering
    \includegraphics{figures/supp-merged-data-model-fits}
    \caption{
        \textbf{Merged datasets bias demographic inferences.}
        The panels represent best-fit models according to 
        different combinations of high- and low- coverage data.
        While statistics from low and high coverage datasets provide consistent
        best-fit demographic models, merged datasets result in biased LD
        statistics (Figure~\ref{fig:supp-merged-data-SEs}). These biased LD
        statistics lead to large downstream biases in inferred demographic
        models. 
    }
    \label{fig:supp-merged-data-fits}
\end{figure}

@@

\subsection{Model choice from likelihood differences}
\label{sec:refit-simulated-data}

The set of models we tested with the full dataset have varying complexity,
ranging from a single origin expansion to multiple stem populations connected
by migration and gene flow. Thus, our models have varying numbers of parameters
($21$ in the simplest model, and $30$ in the most heavily parameterized). Our
analyses found that the more complex models (multiple stems with ongoing
migration) provided the best fit to the data. To ensure that these more complex
models are not simply fitting noise in our data, we simulated data under both
simpler and more complex models, with same sample sizes and similar genome
length to that used in this study. From these, we fit each of our models and
checked that the correct model was chosen.

While simulations of datasets of this size are rapid, computing LD statistics
is computationally burdensome. We simulated 500 1Mb regions under our best fit
single origin, continuous migration, and merger with stem migration models.
Using these simulated regions, we constructed 500 bootstrap replicate datasets
by sampling with replacement. Each of these 500 replicate sets were fit using
the models reported in the main text, and we compared log-likelihoods across
each fit (Figure~\ref{fig:refit-lls}). In each case, even in the simpler single
origin simulations, the correct model was recovered.  We conclude that our
inferences are not simply fitting statistical noise, and the observed
differences in log-likelihoods indicate a better recapitulation of the true
history.

\begin{table}[ht]
    \caption{
        \label{tab:confusion-matrix}
        \textbf{Something like a confusion matrix}
        \aprcomment{fix/fill caption}
    }
    \centering
    \begin{tabular}[t]{lcccc}
        \toprule
        Simulated model & Single origin & Cont. migration & Merger w/out stem mig. & Merger w/stem mig. \\
        \midrule
        Single origin & 494 & 3 & 1 & 2 \\
        Cont. migration & 0 & 500 & 0 & 0 \\
        Merger w/stem mig. & 0 & 0 & 1 & 499 \\
        \bottomrule
    \end{tabular}
\end{table}

The single origin model is nested within the more complex models (with
migration from the additional stem population fixed to zero), those more
complex models can always recover the log-likelihood found by the single origin
model. We therefore explored the broader parameter space that included gene
flow from the additional stem population. In these fits, the optimization
converged either back to the nested single origin model, or it found a local
optimum with lower log-likelihood than the simpler model (see first panel in
Figure~\ref{fig:refit-lls}).

\begin{figure}[htb]
    \centering
    \includegraphics{./figures/supp-refit-lls.pdf}
    \caption{
        \textbf{Fits of four models to simulated data.} In three scenarios, a
        simple single origin model and our preferred continuous migration and
        merger with stem migration models, we simulated 500 replicates of 500
        Mb with the same populations and sample sizes from our reported best
        fit models. We then fit the four models reported in the main text to
        each of these simulated datasets and compared log-likelihoods. In each
        case, the simulated model is correctly chosen, even under the single
        origin scenario. We conclude our more complex models (with more
        parameters) are not simply fitting statistical noise.
    }
    \label{fig:refit-lls}
\end{figure}
